Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass

### 2025-01-02 paper/paper.tex
saving work

\title{Space for WHAM: a multi-region, multi-stock generalization of the
Woods Hole Assessment Model with an application to black sea bass}
\author{Timothy J. Miller\textsuperscript{1} \and Kierstin
Curti \and Alex Hansell}
\date{14 November, 2024}

@@

A state-space statistical approach and maximum marginal likelihood or
Bayesian fitting of stock assessment models allows estimation of time-
and age-varying population attributes as random effects Miller et al.
(2016b). This estimation approach is considered an essential feature of
gold-standard assessment models that we use in tactical management of
commercially important fish stocks (Punt et al. 2020). The State-space
Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed
and remains widely used within ICES to assess European fish stocks.
Various state-space models are being used to manage cod and plaice
stocks in the waters of Eastern Canada (Perreault et al. 2020; Varkey et
al. 2022) and to the south, the Woods Hole Assessment Model (WHAM, Stock
and Miller 2021) is now used to assess many fish stocks in the Northwest
Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).

WHAM is an R package developed and maintained at NOAA's Northeast
Fisheries Science Center (\url{https://timjmiller.github.io/wham},
Miller and Stock 2020; Stock and Miller 2021). WHAM can be configured to
fit a wide range of age-structured models from traditional statistical
catch-at-age models without any random effects to models with several
time and age varying process errors and possibly effects of
environmental covariates on various demographic parameters. Like SAM,
WHAM models are built using the Template Model Builder package (TMB,
Kristensen et al. 2016) which provides a computationally efficient means
of fitting an extremely wide class of models with random effects. WHAM
has undergone active development since its creation and includes random
effects options for in recruitment, inter-annual transitions numbers at
age (hereafter referred to as ``survival''), fishery and index
selectivity, natural mortality, and catchability.

However, WHAM has up to now only allowed models with one stock
(component) and a single region and without seasonal changes in stock
dynamics. Using such models for stocks that have subcomponents with
varying seasonal movement can provide incorrect inferences and poor
management advice (Ying et al. 2011; Cao et al. 2014; Bosley et al.
2022). Furthermore, the ability to account for spatial structure and
model multiple stocks are also important features of leading-edge
assessment modeling frameworks (Punt et al. 2020). We describe here the
implementation of these features and other extensions since Stock and
Miller (2021) in WHAM version 2.0. Many of these new configuration
options can be useful whether modeling 1 or more stocks and regions.

This extension of WHAM was developed in concert with a stock assessment
for black sea bass through the NEFSC research track assessment process
where new modeling frameworks were recently examined and modeling
multiple stocks or stock components simultaneously was of interest
(NEFSC 2023). \textcolor{red}{[Text on black sea bass here]} We apply
WHAM 2.0 to two stock components of black sea bass off the coast of the
NEUS and evaluate evidence for alternative hypotheses of temporal
variation and effects specifically of bottom temperature on recruitment
and natural mortality of age 1 individuals.

@@

Many of the options and equations of WHAM version 2.0 are the same as
those in Stock and Miller (2021), so we will only describe extensions
and differences that have occurred since their first description of
WHAM. The new version of WHAM can model multiple stocks and survival,
movement, harvest and natural mortality are tracked for each stock. Much
of the description below is for a specific stock \(s\), but, for
simplicity, this subscript is implicit except when necessary.

@@

As in previous versions of WHAM, the transitions in numbers at age from
one year to another after recruitment can be treated deterministically
or as functions of random effects. The predicted numbers at age in year
\(y\) at age \(a\) for a given stock are vector analogs
(\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the
standard WHAM model (Stock and Miller 2021). For ages
\(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected
number alive in each region at the beginning of the following year and
next age class age can be obtained from the first \(n_R\) elements of
the vector \[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\] The numbers
alive in each region can also be modeled more simply using the
sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions
in abundance at age is \begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*} for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}.
\end{equation*} When the transitions in abundance at age are treated
deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). The stock-
and region-specific errors \(\boldsymbol{\varepsilon}_{y,a}\) are
independent, but the same correlation structures as previous versions
are possible across ages and years for a given stock and region. When
there is autocorrelation with age, WHAM now assumes this applies only to
ages \(a>1\) by default so that recruitment random effects are
independent of those for for the annual transitions of older age
classes. So the general covariance structure for a given stock at ages
\(a>1\) in region \(r\) is that of a two-dimensional first-order
autoregressive (2DAR1) process \begin{equation*}
  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)} 
\end{equation*} and that for age 1 is just AR1 across years
\begin{equation*}
  Cov\left(\epsilon_{y,1},\epsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
\end{equation*} Since recruitment for a given stock currently only
occurs in one region \(r_s\) there is only a single time-varying
recruitment random effect for each stock.

@@

For each season, there are at most \(n_R-1\) parameters determining
movement among regions given starting an the season in region \(r\) in
either the sequential or simultaneous configurations. Movement
parameters are estimated on a transformed scale via a link function
\(g(\cdot)\). If survival and movement are occur simultaneously, the
parameters are estimated a log link function is used and if they are
separable, an additive logit link function (like a multinomial
regression) is used. On the transformed scale, the general model for the
movement parameter from region \(r\) to \(r'\) in season \(i\) and year
\(y\) for individuals of age \(a\) is a linear function of both random
and environmental effects: \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
\end{equation*} The random effects \(\epsilon_{r\rightarrow r',y,a,i}\)
are season-, and(or) region-to-region-specific and modeled most
generally as 2DAR1 random effects with age and(or) year where the
covariance is \begin{equation*}
  Cov\left(\epsilon_{r\rightarrow r',y,a,i},\epsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
\end{equation*} similar to how WHAM models variation in survival,
natural mortality, and selectivity. Effects of covariate \(E_k\) can be
age-, season-, and(or) region-to-region-specific
\(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial
options in the previous versions of WHAM for effects on recruitment and
natural mortality are available.

There is currently no likelihood component for tagging data. Therefore,
movement parameters would generally either need to be fixed or assumed
to have some prior distribution, possibly based on external parameter
estimates. We include prior distributions for the season and
region-to-region specific (mean) movement parameters which are treated
as random effects with the mean defined by the initial value of the
fixed effect counterpart and standard deviation \begin{equation*}
  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
  \end{equation*} When priors are used, the movement is defined instead
as \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
  \end{equation*}

@@

Natural mortality options have been expanded in WHAM. When not
estimated, (mean) mortality rates may be stock-, region-, and
age-specific. When random effects are used, the same 2DAR1 structure
with age and year as described by Stock and Miller (2021) can be
configured for a given stock and region. Any environmental covariate
effects can be stock-, region-, and age-specific. So the general model
for natural mortality is \begin{equation*}
  \log M_{y,a,r} = \theta_{M,r} + \epsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
\end{equation*} The general covariance structure for random effects are
modeled most generally as 2DAR1 random effect with age and(or) year
where the covariance is \begin{equation*}
  Cov\left(\epsilon_{M,y,a,r},\epsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
\end{equation*}

@@

Catchability of the index can also be treated as functions of normal
random effects and(or) environmental covariate effects
\[log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}\]
where \(u_{m}\) and \(l_{m}\) are the upper and lower bounds of
catchability for index \(m\) (defaults are 0 and 1000) and the general
covariance structure for the annual random effects is AR1
\[Cov\left(\epsilon_{q,m,y},\epsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.\]

@@

For \(X\%\) SPR-based reference points, we use a Newton method and
iterate \begin{equation}\label{eq:newton}
  \log\widetilde{F}^{(i)} = \log\widetilde{F}^{(i-1)} - \frac{g\left(\log\widetilde{F}^{(i-1)}\right)}{g'\left(\log\widetilde{F}^{(i-1)}\right)}
\end{equation} where \(g(\log F)\) is the difference between the
weighted sums of spawning biomass per recruit at \(F\) and \(X\)\% of
unfished spawning biomass per recruit across stocks: \begin{equation*}
  g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
\end{equation*} where \(\upphi_s\left(F=0\right)\) is the equilibrium
unfished spawning biomass per recruit. \(g'(\log F)\) is the derivative
of \(g\) with respect to \(\log F\), and the weights to use for each
stock \(\lambda_s\) can be specified by the user or relative to the
average of recruitment for each stock over the same years the user
defines to calculate ``static'\,' equilibrium spawning biomass and
yield.

@@

@@

Leading up to the 2023 peer-reviewed assessment, a working group
(hereafter referred to as ``working group'') composed of scientists from
federal, state, and academic institutions determined an optimal data and
model configuration for the black sea bass stock using the multi-stock
and multi-region extension of WHAM described above (NEFSC 2023). This
assessment included the spatial features and investigated inclusion of
hypothesized environmental drivers that were prioritized research
recommendations from previous black sea bass assessments.

Below we describe the assumptions and configuration of the assessment
model as determined by the working group as well as the alternative
assumptions for recruitment and natural mortality in the models we fit
to evaluate alternative hypotheses of bottom temperature effects on
black sea bass.

@@

The two monthly movement matrices are \begin{equation*}
\boldsymbol{\mu}_{1} = 
  \begin{bmatrix}
     1-\mu_{\text{N}\rightarrow \text{S}} & \mu_{\text{N}\rightarrow \text{S}} \\
     0 & 1 \\
  \end{bmatrix}
\end{equation*} for the portion of the year after spawning and
\begin{equation*}
\boldsymbol{\mu}_{2} = 
  \begin{bmatrix}
     1 &  0 \\
     \mu_{\text{S}\rightarrow \text{N}} & 1-\mu_{\text{S}\rightarrow \text{N}} \\
  \end{bmatrix}
\end{equation*} for the portion of the year before spawning. As noted in
the description of the general WHAM model, tagging data are not yet
allowed. However, the working group also fit a Stock Synthesis model
(Methot and Wetzel 2013) which provided estimated movement rate
parameters and standard errors that were used to configure the priors
for WHAM (see Supplementary Materials).

@@

\hypertarget{uncertainty-recreational-cpa-index-observations}{%
\subsection*{Uncertainty recreational CPA index
observations}\label{uncertainty-recreational-cpa-index-observations}}
\addcontentsline{toc}{subsection}{Uncertainty recreational CPA index
observations}

The estimated coefficients of variation (CVs) provided by the analyses
to generate the recreational catch per angler (recreational CPA) index
ranged between 0.02 and 0.06 which the working group felt did not
capture the true observation uncertainty in the index with regard to its
relationship to stock abundance. In many of the initial runs as well as
the base model we allowed a scalar multiple of the standard deviation of
the log aggregate index to be estimated for these indices in the
northern and southern regions. Models that successfully estimated these
scalars indicated standard deviations for these surveys to be
approximately 5 times the input value and this value was fixed in many
preliminary runs to avoid dealing with convergence problems. However,
the base model successfully estimated these scalars. The model estimates
are negligibly affected by estimating these scalars, but we felt
estimating these parameters allowed uncertainty in model output to be
more properly conveyed.

@@

\begin{equation}\label{eq:expected-recruitment}
E\left(\log N_{y,1}|x_y\right) = \mu_{R} + \beta_{R} x_y.
\end{equation} Similarly, expected log-natural morality as a function of
bottom temperature is \begin{equation}\label{eq:expected-M1}
E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y
\end{equation} Because age 1 fish for the northern component can exist
in both regions after January 1, natural mortality is acting in both
regions for this stock. For models with covariate and/or annual random
effects for age 1 fish we assume them only in the northern regions for
the northern component. The corresponding random effects are
\begin{equation}\label{eq:Rec-re}
\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \epsilon_{y,1}
\end{equation} and \begin{equation}\label{eq:M-re}
\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \epsilon_{M,y}.
\end{equation} The constant or mean log-natural mortality rate is
assumed to be \(\mu_{M,a} = \log(0.4)\) for all ages as recommended by
the working group. Because the bottom temperature anomalies and random
effects are centered at 0, mean log-natural mortality at age 1 over the
time series should be approximately equal to \(\mu_{M,1}\) for models
that include those effects.

\hypertarget{model-fitting-diagnostics-and-projections}{%
\subsection*{Model fitting, diagnostics, and
projections}\label{model-fitting-diagnostics-and-projections}}
\addcontentsline{toc}{subsection}{Model fitting, diagnostics, and
projections}

@@

To illustrate projections with environmental effects we projected the
best black sea bass model under three alternative scenarios where the
AR1 time-series model for bottom temperature continues into the
projection years, where we project with bottom temperature being the
average of the most recent 5 years, and where bottom temperature
increases following a predictions from a simple linear regression of the
estimated bottom temperature anomalies over time.

@@

>>>>

\title{Space for WHAM: a multi-region, multi-stock generalization of the
Woods Hole Assessment Model with an application to black sea bass}
\author{Timothy J. Miller\textsuperscript{1} \and Kierstin
Curti \and Alex Hansell}
\date{02 January, 2025}

@@

A state-space statistical approach and maximum marginal likelihood or
Bayesian fitting of stock assessment models allows estimation of time-
and age-varying population attributes as random effects Miller et al.
(2016b). This estimation approach is considered an essential feature of
gold-standard assessment models that we use in tactical management of
commercially important fish stocks (Punt et al. 2020). The State-space
Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed
and remains widely used within ICES to assess European fish stocks.
Various state-space models are being used to manage cod and plaice
stocks in the waters of Eastern Canada (Perreault et al. 2020; Varkey et
al. 2022) and to the south, the Woods Hole Assessment Model (WHAM,
Miller and Stock 2020; Stock and Miller 2021) is now used to assess many
fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC
2024).

WHAM is an R package developed and maintained at NOAA's Northeast
Fisheries Science Center {[}\url{https://timjmiller.github.io/wham}{]}.
WHAM can be configured to fit a wide range of age-structured models from
traditional statistical catch-at-age models without any random effects
to models with several time and age varying process errors and effects
of environmental covariates on various population parameters. Like SAM,
WHAM models are built using the Template Model Builder package (TMB,
Kristensen et al. 2016) which provides a computationally efficient means
of fitting an extremely wide class of models with random effects. WHAM
has undergone active development since its creation and includes random
effects options for recruitment, inter-annual transitions of numbers at
age (hereafter referred to as ``survival''), fishery and index
selectivity, natural mortality, and catchability.

However, WHAM has only allowed one stock (component) and a single region
and without seasonal changes in stock dynamics. Using such models for
stocks that have subcomponents with varying seasonal movement can
provide incorrect inferences and poor management advice (Ying et al.
2011; Cao et al. 2014; Bosley et al. 2022). Furthermore, the ability to
account for spatial structure and model multiple stocks are also
important features of leading-edge assessment modeling frameworks (Punt
et al. 2020). We describe here the implementation of these features and
other extensions since Stock and Miller (2021) in WHAM version 2.0. Many
of these new configuration options can also be useful when modeling a
single stock and region.

We developed this extension of WHAM in concert with a research track
assessment process for black sea bass where new modeling frameworks were
recently examined and modeling two stock components simultaneously and
temporal changes in distribution was of interest (NEFSC 2023). To
illustrate the usage of WHAM 2.0 we apply it to two stock components of
black sea bass off the coast of the NEUS using most of the model
assumptions that were accepted during the peer-review process for the
black sea bass assessment. However, here we evaluate evidence for
alternative hypotheses of temporal variation and effects specifically of
bottom temperature on recruitment and natural mortality of age 1
individuals.

When modeling multiple stock components occurring in different regions
further dimensionality is added to the class of possible equilbrium
reference points (e.g., Kapur et al. 2021). Short term projections are
also an important part of the management process and the separation of
observation and process errors in state-space models allows uncertainty
in projected attributes such as recruitment and SSB without external
simulation exercises. Finally, the inclusion of explicit effects of
environmental covariates in the model also requires some assumptions on
how they are treated in projections and reference points. We demonstrate
some of options of WHAM version 2.0 provide biological reference points
and short-term projections.

@@

Many of the options and equations of WHAM version 2.0 are the same as
those in Stock and Miller (2021), so we will only describe extensions
and differences that have occurred since their first description of
WHAM. The new version of WHAM can model multiple stocks each with their
own movement, harvest, and natural mortality. Much of the description
below is for a specific stock \(s\), but, for simplicity, this subscript
is implicit except when necessary.

@@

As in previous versions of WHAM, the transitions in numbers at age from
one year to another after recruitment can be treated deterministically
or as functions of random effects. The predicted numbers at age in year
\(y\) at age \(a\) for a given stock are vector analogs
(\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the
standard WHAM model (Stock and Miller 2021). For ages
\(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected
number alive in each region at the beginning of the following year and
next age class age can be obtained from the first \(n_R\) elements of
the vector \[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\] The numbers
alive in each region can also be modeled more simply using the
sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions
in abundance at age is \begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*} for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
\end{equation*} where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector
of region-specific errors for a given stock, which are independent
across stocks and regions, but the same correlation structures as
previous versions are possible across ages and years for a given stock
and region. When there is autocorrelation with age, WHAM now assumes
this applies only to ages \(a>1\) by default so that recruitment random
effects are independent of those for for the annual transitions of older
age classes. So the general covariance structure for a given stock at
ages \(a>1\) in region \(r\) is that of a two-dimensional first-order
autoregressive (2DAR1) process \begin{equation*}
  Cov\left(\varepsilon_{y,a,r},\varepsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)} 
\end{equation*} and that for age 1 is just AR1 across years
\begin{equation*}
  Cov\left(\varepsilon_{y,1},\varepsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
\end{equation*} When the transitions in abundance at age are treated
deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). Since
recruitment for a given stock currently only occurs in one region
\(r_s\) there is only a single time-varying recruitment random effect
for each stock.

@@

For each season, there are at most \(n_R-1\) parameters determining
movement among regions given starting an the season in region \(r\) in
either the sequential or simultaneous configurations. Movement
parameters are estimated on a transformed scale via a link function
\(g(\cdot)\). If survival and movement are occur simultaneously, the
parameters are estimated a log link function is used and if they are
separable, an additive logit link function (like a multinomial
regression) is used. On the transformed scale, the general model for the
movement parameter from region \(r\) to \(r'\) in season \(i\) and year
\(y\) for individuals of age \(a\) is a linear function of both random
and environmental effects: \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
\end{equation*} The random effects
\(\varepsilon_{r\rightarrow r',y,a,i}\) are season-, and(or)
region-to-region-specific and modeled most generally as 2DAR1 random
effects with age and(or) year where the covariance is \begin{equation*}
  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
\end{equation*} similar to how WHAM models variation in survival,
natural mortality, and selectivity. Effects of covariate \(E_k\) can be
age-, season-, and(or) region-to-region-specific
\(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial
options in the previous versions of WHAM for effects on recruitment and
natural mortality are available.

There is currently no likelihood component for tagging data. Therefore,
movement parameters would generally either need to be fixed or assumed
to have some prior distribution, possibly based on external parameter
estimates. We include prior distributions for the season and
region-to-region specific (mean) movement parameters which are treated
as random effects with the mean defined by the initial value of the
fixed effect counterpart and standard deviation \begin{equation*}
  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
  \end{equation*} When priors are used, the movement is defined instead
as \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
  \end{equation*}

@@

Natural mortality options have been expanded in WHAM. When not
estimated, (mean) mortality rates may be stock-, region-, and
age-specific. When random effects are used, the same 2DAR1 structure
with age and year as described by Stock and Miller (2021) can be
configured for a given stock and region. Any environmental covariate
effects can be stock-, region-, and age-specific. So the general model
for natural mortality is \begin{equation*}
  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
\end{equation*} The general covariance structure for random effects are
modeled most generally as 2DAR1 random effect with age and(or) year
where the covariance is \begin{equation*}
  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
\end{equation*}

@@

Catchability of the index can also be treated as functions of normal
random effects and(or) environmental covariate effects
\[log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}\]
where \(u_{m}\) and \(l_{m}\) are the upper and lower bounds of
catchability for index \(m\) (defaults are 0 and 1000) and the general
covariance structure for the annual random effects is AR1
\[Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.\]

@@

For \(X\%\) SPR-based reference points, we use a Newton method and
iterate \begin{equation}\label{eq:newton}
  \log\widetilde{F}^{(i)} = \log\widetilde{F}^{(i-1)} - \frac{g\left(\log\widetilde{F}^{(i-1)}\right)}{g'\left(\log\widetilde{F}^{(i-1)}\right)}
\end{equation} where \(g(\log F)\) is the difference between the
weighted sums of spawning biomass per recruit at \(F\) and \(X\)\% of
unfished spawning biomass per recruit across stocks:
\begin{equation}\label{eq:newton-obj}
  g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
\end{equation} where \(\upphi_s\left(F=0\right)\) is the equilibrium
unfished spawning biomass per recruit. \(g'(\log F)\) is the derivative
of \(g\) with respect to \(\log F\), and the weights to use for each
stock \(\lambda_s\) can be specified by the user or relative to the
average of recruitment for each stock over the same years the user
defines to calculate ``static'\,' equilibrium spawning biomass and
yield.

@@

Black sea bass are a temperate reef fish in the western Atlantic Ocean
ranging from the Gulf of Mexico to the Gulf of Maine. Fish north of Cape
Hatteras, NC, are considered to comprise a single stock unit, but
individual populations exhibit spawning site fidelity (Able and Stanton
Hales, Jr. 1997; Fabrizio et al. 2013). Fish in this northern stock unit
perform seasonal migrations out on the continental shelf in the fall and
back to their inshore spawning areas during the spring (Musick and
Mercer 1977). Analyses of tagging studies have demonstrated that the
extent of seasonal migration varies along the coast such that fish from
populations off of southern New England and further north move offshore
and as far south as the coasts of Virgina and North Carolina, whereas
fish in the southern portion of the stock range generally move shorter
distances between inshore and offshore areas of similar latitude (Moser
and Shepherd 2009).

@@

Leading up to the 2023 peer-reviewed assessment, a working group
(hereafter referred to as ``Working Group'') composed of scientists from
federal, state, and academic institutions determined an optimal data and
model configuration for the black sea bass stock using the multi-stock
and multi-region extension of WHAM described above (NEFSC 2023). This
assessment included the spatial features and investigated inclusion of
hypothesized environmental drivers that were prioritized research
recommendations from previous black sea bass assessments.

Below we describe the assumptions and configuration of the assessment
model as determined by the Working Group as well as the alternative
assumptions for recruitment and natural mortality in the models we fit
to evaluate alternative hypotheses of bottom temperature effects on
black sea bass.

@@

The two monthly movement matrices are \begin{equation*}
\boldsymbol{\mu}_{1} = 
  \begin{bmatrix}
     1-\mu_{\text{N}\rightarrow \text{S}} & \mu_{\text{N}\rightarrow \text{S}} \\
     0 & 1 \\
  \end{bmatrix}
\end{equation*} for the portion of the year after spawning and
\begin{equation*}
\boldsymbol{\mu}_{2} = 
  \begin{bmatrix}
     1 &  0 \\
     \mu_{\text{S}\rightarrow \text{N}} & 1-\mu_{\text{S}\rightarrow \text{N}} \\
  \end{bmatrix}
\end{equation*} for the portion of the year before spawning. As noted in
the description of the general WHAM model, tagging data are not yet
allowed. However, the Working Group also fit a Stock Synthesis model
(Methot and Wetzel 2013) which provided estimated movement rate
parameters and standard errors that were used to configure the priors
for WHAM (see Supplementary Materials).

@@

\hypertarget{uncertainty-in-recreational-cpa-index-observations}{%
\subsection*{Uncertainty in recreational CPA index
observations}\label{uncertainty-in-recreational-cpa-index-observations}}
\addcontentsline{toc}{subsection}{Uncertainty in recreational CPA index
observations}

The estimated coefficients of variation (CVs) provided by the analyses
to generate the recreational catch per angler (recreational CPA) index
ranged between 0.02 and 0.06 which the Working Group felt did not
capture the true observation uncertainty in the index with regard to its
relationship to stock abundance. In many of the initial runs as well as
the base model we allowed a scalar multiple of the standard deviation of
the log aggregate index to be estimated for these indices in the
northern and southern regions. Models that successfully estimated these
scalars indicated standard deviations for these surveys to be
approximately 5 times the input value and this value was fixed in many
preliminary runs to avoid dealing with convergence problems. However,
the base model successfully estimated these scalars. The model estimates
are negligibly affected by estimating these scalars, but we felt
estimating these parameters allowed uncertainty in model output to be
more properly conveyed.

@@

\begin{equation}\label{eq:expected-recruitment}
E\left(\log N_{y,1}|x_y\right) = \mu_{R} + \beta_{R} x_y.
\end{equation} Similarly, expected log-natural morality as a function of
bottom temperature is \begin{equation}\label{eq:expected-M1}
E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y
\end{equation} Because age 1 fish for the northern component can exist
in both regions after January 1, natural mortality is acting in both
regions for this stock. For models with covariate and/or annual random
effects for age 1 fish we assume them only in the northern regions for
the northern component. The corresponding random effects are
\begin{equation}\label{eq:Rec-re}
\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \varepsilon_{y,1}
\end{equation} and \begin{equation}\label{eq:M-re}
\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \varepsilon_{M,y}.
\end{equation} The constant or mean log-natural mortality rate is
assumed to be \(\mu_{M,a} = \log(0.4)\) for all ages as recommended by
the Working Group. Because the bottom temperature anomalies and random
effects are centered at 0, mean log-natural mortality at age 1 over the
time series should be approximately equal to \(\mu_{M,1}\) for models
that include those effects.

\hypertarget{model-fitting-diagnostics-brps-and-projections}{%
\subsection*{Model fitting, diagnostics, BRPs, and
projections}\label{model-fitting-diagnostics-brps-and-projections}}
\addcontentsline{toc}{subsection}{Model fitting, diagnostics, BRPs, and
projections}

@@

We produced SPR-based BRP estimates (\(F_{40\%}\) and SSB\(_{40\%}\))
from the best performing model using a variety of options provided in
WHAM. We generated annual BRP estimates based on the annual weight at
age, maturity at age, M, fleet selectivity, and either the annual
recruitment random effect (Eq. \ref{eq:Rec-re}) or the annual expected
recruitment (Eq. \ref{eq:expected-recruitment}). We also calculated
status as the ratios of annual average F and SSB to the corresponding
reference points. Finally, we calculated ``prevailing'' BRPs using the
average weight at age, maturity at age, M, fleet selectivity from the
last 5 years (2017 to 2021) and average recruitment over years 2000 to
2021, and made Kobe plots of joint status (Kobe plots) for the terminal
year of the model (2021). All BRPs and status estimates are generated
for each region as well as for the total stock.

We projected the best black sea bass model under three alternative
scenarios for the bottom temperature covariate where 1) the AR1
time-series model continues into the projection years, 2) the average of
the most recent 5 years is projected, and 3) bottom temperature
increases in projection years following a prediction from a simple
linear regression of the estimated bottom temperature anomalies over
time. We specified fishing mortality in projection years to be constant
at the value in the last model year (2021). Average weight, natural
mortality, movement, and maturity at age are used in projection years
whereas the AR1 processes for all numbers at age random effects are
continued.

@@

In our investigation of bottom temperature effects on recruitment and
natural mortality and time-varying random effects on natural mortality
at age 1, we found only including effects on recruitment for the
northern stock component to be the best model with respect to AIC. Black
sea bass in the NEUS is at the northern extent of its range and the
estimated increase in expected recruitment with temperature for the
northern stock component is consistent with its range extension to the
north. Opposite effects of temperature on recruitment or population size
would be expected for species in the same general area that are at the
southern extent of their range (Gabriel 1992). For example, higher
recruitment of the Southern New England-Mid-Atlantic stock of yellowtail
flounder is correlated with more cold water persistence into summer fall
on the Northeast US shelf Miller et al. (2016b).
